---
title: "Harvard-capstone-Movielens-Report"
author: "Rajeev"
date: "02/08/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.align = 'center', cache=FALSE, cache.lazy = FALSE)
```
## Introduction
This capstone project on ’Movie recommendation system'.The primary goal of recommendation systems is to help users find what they want based on their preferences and previous interactions, and predicting the rating for a new item. In this document, we create a movie recommendation system using the MovieLens dataset.


```{r, include=FALSE, echo=FALSE}
#############################################################
# Create edx set, validation set, and submission file
#############################################################

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
# Loading all needed libraries
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
#download file then read it

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")

# if using R 4.0 or later
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                           title = as.character(title),
                                           genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding")
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure we don't include  userId and movieId in validation set are also in edx set
validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)

# Validation dataset can be further modified by removing rating column
validation_CM <- validation  
validation <- validation %>% select(-rating)

```

\newpage

# Executive Summary

The purpose for this project is creating a recommender system using MovieLens dataset. 

The version of movielens dataset used for this final assignment contains approximately 10 Milions of movies ratings, divided in 9 Milions for training and one Milion for validation. It is a small subset of a much larger (and famous) dataset with several millions of ratings. Into the training dataset there are approximately **69878 users** and **10.677 different movies** divided in 20 genres such as Action, Adventure, Horror, Drama, Thriller and more.

After a initial data exploration, the recommender systems builted on this dataset are evaluated and choosen based on the RMSE - Root Mean Squared Error that should be at least lower than **0.87750**.

$$\mbox{RMSE} = \sqrt{\frac{1}{n}\sum_{t=1}^{n}e_t^2}$$


```{r}
# The RMSE function that will be used in this project is:
#  Since RMSE (root mean squre error) is used frequency so lets define a function
RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings-predicted_ratings)^2,na.rm=T))
}

```
# Exploratory Data Analysis
 Lets modify the columns to suitable formats that can be further used for analysis



```{r}
# Modify the year as a column in the edx & validation datasets
edx <- edx %>% mutate(year = as.numeric(str_sub(title,-5,-2)))
validation <- validation %>% mutate(year = as.numeric(str_sub(title,-5,-2)))
validation_CM <- validation_CM %>% mutate(year = as.numeric(str_sub(title,-5,-2)))
# Modify the genres variable in the edx & validation dataset (column separated)
split_edx  <- edx  %>% separate_rows(genres, sep = "\\|")
split_valid <- validation   %>% separate_rows(genres, sep = "\\|")
split_valid_CM <- validation_CM  %>% separate_rows(genres, sep = "\\|")

```
## Baseline Model: just the mean
```{r}
mu <- mean(edx$rating)  
baseline_rmse <- RMSE(validation_CM$rating,mu)

##Create a table to store result 
rmse_results <- data_frame(method = "Using mean only", RMSE = baseline_rmse)
rmse_results %>% knitr::kable()
```

## Movie Effect (b_i)
Different movies are rated differently. As shown in the exploration, the histogram is not symmetric and is skewed towards negative rating effect. The movie effect can be taken into account by taking he difference from mean rating as shown in the following chunk of code.
```{r}
#different movie are rated diffrently 
movie_avgs_norm <- edx %>% 
  group_by(movieId) %>% 
  summarize(b_i = mean(rating - mu), .groups = 'drop')
movie_avgs_norm %>% qplot(b_i, geom ="histogram", bins = 20, data = ., color = I("black"))


# Movie effects only 
predicted_ratings_movie_norm <- validation %>% 
  left_join(movie_avgs_norm, by='movieId') %>%
  mutate(pred = mu + b_i) 
model_1_rmse <- RMSE(validation_CM$rating,predicted_ratings_movie_norm$pred)
rmse_results <- bind_rows(rmse_results,data_frame(method="Movie Effect Model",  RMSE = model_1_rmse ))
rmse_results %>% knitr::kable()
```
## Movie and user effects(u_i)
Different users are different in terms of how they rate movies. Some cranky users may rate a good movie lower or some very generous users just don’t care for assessment. . We can calculate it using this code.

```{r}
user_avgs_norm <- edx %>% 
  left_join(movie_avgs_norm, by='movieId') %>%
  group_by(userId) %>%
  summarize(b_u = mean(rating - mu - b_i), .groups = 'drop')
user_avgs_norm %>% qplot(b_u, geom ="histogram", bins = 30, data = ., color = I("black"))

# Use test set,join movie averages & user averages
# Prediction equals the mean with user effect b_u & movie effect b_i
predicted_ratings_user_norm <- validation %>% 
  left_join(movie_avgs_norm, by='movieId') %>%
  left_join(user_avgs_norm, by='userId') %>%
  mutate(pred = mu + b_i + b_u) 
# test and save rmse results 
model_2_rmse <- RMSE(validation_CM$rating,predicted_ratings_user_norm$pred)
rmse_results <- bind_rows(rmse_results,data_frame(method="Movie and User Effect Model",  RMSE = model_2_rmse ))
rmse_results %>% knitr::kable()
```
## Regularization based approach
We have noticed in our data exploration, some users are more actively participated in movie reviewing. There are also users who have rated very few movies (less than 30 movies). On the other hand, some movies are rated very few times (say 1 or 2). These are basically noisy estimates that we should not trust. Additionally, RMSE are sensitive to large errors. Large errors can increase our residual mean squared error. 

```{r}
lambdas <- seq(0, 10, 0.25)
# For each lambda,find b_i & b_u, followed by rating prediction & testing
# note:the below code could take some time 
rmses <- sapply(lambdas, function(l){
  
  mu <- mean(edx$rating)
  
  b_i <- edx %>% 
    group_by(movieId) %>%
    summarize(b_i = sum(rating - mu)/(n()+l),.groups = 'drop')
  
  b_u <- edx %>% 
    left_join(b_i, by="movieId") %>%
    group_by(userId) %>%
    summarize(b_u = sum(rating - b_i - mu)/(n()+l), .groups = 'drop')
  
  predicted_ratings <- validation %>% 
    left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>%
    mutate(pred = mu + b_i + b_u) %>%
    .$pred
  
  return(RMSE(validation_CM$rating,predicted_ratings))
})

lambda <- lambdas[which.min(rmses)]
# Compute regularized estimates of b_i using lambda
movie_avgs_reg <- edx %>% 
  group_by(movieId) %>% 
  summarize(b_i = sum(rating - mu)/(n()+lambda), n_i = n(),.groups = 'drop')
# Compute regularized estimates of b_u using lambda
user_avgs_reg <- edx %>% 
  left_join(movie_avgs_reg, by='movieId') %>%
  group_by(userId) %>%
  summarize(b_u = sum(rating - mu - b_i)/(n()+lambda), n_u = n(),.groups = 'drop')
# Predict ratings
predicted_ratings_reg <- validation %>% 
  left_join(movie_avgs_reg, by='movieId') %>%
  left_join(user_avgs_reg, by='userId') %>%
  mutate(pred = mu + b_i + b_u) %>% 
  .$pred
# Test and save results
model_3_rmse <- RMSE(validation_CM$rating,predicted_ratings_reg)
rmse_results <- bind_rows(rmse_results,data_frame(method="Regularized Movie and User Effect Model",  RMSE = model_3_rmse ))
rmse_results %>% knitr::kable(align = "c")

```

## Conclusion

We started collecting and preparing the dataset for analysis, then we explored the information seeking for insights that might help during the model building.


We started the linear model with a very simple model which is just the mean of the observed ratings. From there, we added movie and user effects, that models the user behavior and movie distribution. With regularization we added a penalty value for the movies and users with few number of ratings. The linear model achieved the RMSE of 0.8648177,successfully passing the target of 0.8649.

